---
title: 'Storm Data Analysis: An exploration of severe weather events from 1950 to
  2011'
author: "LarionovaAnna"
date: "Sunday, May 24, 2015"
output: html_document
---
## Synopsis
Storms have the ability to cause devistation to both human well-being and property. Damage caused by storms can be viewed along two metrics. Cost to human life, and cost of property damage. In this paper, we look at data from the National Weather Service and use it to examine the costs to the US for various weather events.
## Data Processing

###switch on library
```{r}
library(ggplot2)
library(dplyr)
source("C:/Documents and Settings/Администратор/RepData_PeerAssessment2/multiplot.R")
```
###take data
```{r}
URL <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
datafile <- "stormData.csv.bz2"
download.file(URL, datafile)
stormdata <- read.csv(bzfile("stormData.csv.bz2"), strip.white = TRUE)
```
###take only USA states
```{r}
stormdata <- stormdata[stormdata$STATE %in% state.abb,]
```
###take data to analyze consequences to population health
```{r}
stormdata_population_health <- stormdata[,c("BGN_DATE", "EVTYPE", "FATALITIES", "INJURIES")]
```
###take data to analyze consequences to economic
```{r}
stormdata_economic <- subset(stormdata, (CROPDMG > 0 | PROPDMG > 0), select = c(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))
```
###extracting year of event
``````{r}
stormdata_population_health$BGN_DATE <- as.character(stormdata_population_health$BGN_DATE)
testregexp <- regexpr("[0-9]{4}", stormdata_population_health$BGN_DATE)
testmatch <- regmatches(stormdata_population_health$BGN_DATE, testregexp)
stormdata_population_health$BGN_DATE <- as.integer(stormdata_population_health$BGN_DATE <- testmatch)
```
###group by year and event
```{r}
stormdata_population_health$EVTYPE <- tolower(stormdata_population_health$EVTYPE)
stormdata_economic$EVTYPE <- tolower(stormdata_economic$EVTYPE)
stormdata_economic$PROPDMGEXP <- tolower(stormdata_economic$PROPDMGEXP)
stormdata_economic$CROPDMGEXP <- tolower(stormdata_economic$CROPDMGEXP)
fat_and_inj <- subset(stormdata_population_health, (FATALITIES > 0 | INJURIES > 0), select = c(BGN_DATE, EVTYPE, FATALITIES, INJURIES))
```
###search words to group them into events 
```{r}
type <- vector(mode = "list", length = 6)
names(type) <- c("storm", "extreme_cold", "heat", "in_sea", "landslide", "other")
type[[1]] <- c("storm", "dam break", "gustnado", "tropical depression", "landspout", "tstmw", "burst", "wet", "turbulence", "lig", "thunder", "dust", "wind", "hurricane", "tornado", "funnel cloud", "flood", "rain", "fog", "torndao", "fld", "stream")
type[[2]] <- c("frost", "cool", "chill", "precip","freez", "mix", "glaze", "sleet", "hypothermia", "low temp", "ice", "icy", "snow", "cold", "blizzard", "swirl", "hail", "winter", "wintry")
type[[3]] <- c("heat wave", "smoke", "dry", "warm", "high temp", "drought", "heat", "hyperthermia", "fire")
type[[4]] <- c("drowning", "beach", "surf", "surge", "coast", "rip current", "tide", "seiche", "sea", "ocean", "tsunami", "marine mishap", "marine", "wave", "waterspout", "water")
type[[5]] <- c("landslide", "avalance", "aval", "rockslide", "slide", "land")
type[[6]] <- c("other", "high", "\\?", "urban", "county")
for (t in names(type)) {
    for (word in type[[t]]) {
        fat_and_inj$EVTYPE <- gsub(paste(".*", word, ".*", sep = ""), t, fat_and_inj$EVTYPE)
        stormdata_economic$EVTYPE <- gsub(paste(".*", word, ".*", sep = ""), t, stormdata_economic$EVTYPE)
    }
}
```
###aggregate data by years and types of events and sum injuries and fatalities
```{r}
fat_and_inj_sum <- aggregate(fat_and_inj[,3:4], fat_and_inj[,1:2], sum)
```
###transform economic damage fields
```{r}
prop_seq <- apply(stormdata_economic, 1, function(x) { if (x["PROPDMGEXP"] == "k") { as.numeric(x["PROPDMG"]) * 1000 } else {if (x["PROPDMGEXP"] == "m") {as.numeric(x["PROPDMG"])*10^6} else {if (x["PROPDMGEXP"] == "b") {as.numeric(x["PROPDMG"])*10^9} else {if (x["PROPDMGEXP"] == "h") {as.numeric(x["PROPDMG"])*100}}}}} )
crop_seq <- apply(stormdata_economic, 1, function(x) { if (x["CROPDMGEXP"] == "k") { as.numeric(x["CROPDMG"]) * 1000 } else {if (x["CROPDMGEXP"] == "m") {as.numeric(x["CROPDMG"])*10^6} else {if (x["CROPDMGEXP"] == "b") {as.numeric(x["CROPDMG"])*10^9} else {if (x["CROPDMGEXP"] == "h") {as.numeric(x["CROPDMG"])*100}}}}} )
stormdata_economic$PROPDMG <- as.numeric(paste(prop_seq, sep = " "))
stormdata_economic$CROPDMG <- as.numeric(paste(crop_seq, sep = " "))
stormdata_economic <- stormdata_economic[,c("EVTYPE", "PROPDMG", "CROPDMG")]
```
###aggregate data by event type only
```{r}
stormdata_economic_sum <- aggregate(stormdata_economic[,2:3], stormdata_economic[1], sum, na.rm = TRUE)
```
##Results
The cost to population health of storms can be measured with injuries and fatalities. The economic cost can be measured in the US dollar cost of the damage to property and crops.
###create graphics
```{r}
i <- ggplot(data = fat_and_inj_sum, 
            aes(y = INJURIES, x = BGN_DATE, colour = EVTYPE)) + 
    ggtitle("Injuries by time in every event") + 
    geom_line() +
    geom_point()
f <- ggplot(data = fat_and_inj_sum, 
            aes(y = FATALITIES, x = BGN_DATE, colour = EVTYPE)) + 
    ggtitle("Fatalities by time in every event") + 
    geom_line() +
    geom_point()
g <- multiplot(i, f, cols = 2)
g
p <- ggplot(data = stormdata_economic_sum, 
            aes(x = EVTYPE, y = PROPDMG, fill = PROPDMG)) + 
    ggtitle("Property damage by event") + 
    geom_bar(position = "dodge", alpha = 0.5, stat = "identity") + 
    ylab("Property damage in $") + xlab("Event group")
c <- ggplot(data = stormdata_economic_sum, 
            aes(x = EVTYPE, y = CROPDMG, fill = CROPDMG)) + 
    ggtitle("Crop damage by event") + 
    geom_bar(position = "dodge", alpha = 0.5, stat = "identity") + 
    ylab("Crop damage in $") + xlab("Event group")
g2 <- multiplot(p, c, cols = 2)
g2
```
Most harmful events are caused by storms and related to them events (rain, hail, tornado, strong wind, floods, etc). Storms bring damage to property, infrastructure, crops, storms and related events have the highest rates of fatalities and injures. Storms in USA bring greater damage and impact on human live than severe cold weather or abnormal heat and fires, landslides or events in sea and at cost area put together.
Extreme heat is on the second place of most harmful events, on the third extreme cold. This two events has a greater impact on crops rather than on property damage. Heat and caused by it wildfires kill people almost as much as storm events in 1992-2011 and in the years of 1995, 1999, 2006 abnormal heat and fires took even more lives than storms accroding to the dataset. 